---
title: "Untitled"
author: "Hintz A."
date: "2026-08-17"
output: html_document
---

```{r}
knitr::opts_chunk$set(eval = FALSE)
```

# Introduction

In this workshop, we will introduce some basic concepts of data analysis and data organization in R.


# Packages

Packages provide additional functions and datasets that extend the capabilities of R.

For this workshop, we will use:

- agridat: agricultural datasets;
- tidyverse: a collection of packages for data manipulation, visualization, and analysis.

```{r, message=FALSE, warning=F}
library(agridat)
library(tidyverse)
```

# Exploring the data

## Selecting a Dataset

The agridat package contains many datasets related to agricultural experiments.

To see the datasets available in the package:

```{r}
data(package = "agridat")
```

For this workshop, we will use the sorghum dataset:
```{r}
dat <- agridat::adugna.sorghum

head(dat)
```
  

In practice, data are often stored in Excel or CSV files.

For example, an Excel file can be imported using:

```{r}
# dat <- read_excel("path/to/your_file.xlsx")
```

A CSV file can be imported using:

```{r}
# dat <- read_csv("path/to/your_file.csv")
```


Question: What are the observations in this dataset? What does each row represent?


## Structure of the Data

Before analyzing a dataset, we should understand its structure.

The str() function provides information about:

- the number of observations;
- the variables;
- the type of each variable;
- some example values.

```{r}
str(dat)
```

## Descriptive summary

The summary() function provides a quick descriptive summary of each variable.

```{r}
summary(dat)
```

## Checking missing values

In R, missing values are represented by NA.

We can count the number of missing values in each variable using:

```{r}
colSums(is.na(dat))
```

If a variable contains missing values, we need to consider how they should be handled before performing an analysis.

For example, when calculating a mean, we can use:

```{r}
mean(dat$yield, na.rm = TRUE)
```

The argument na.rm = TRUE tells R to remove missing values before calculating the mean.

## Frequency table

The table() function can be used to count combinations of categorical variables.

For example, we can examine how many observations are available for each combination of location and year:

```{r}
table(dat$loc, dat$year)
```


# Pipes

The pipe allows us to pass the result of one operation directly into another operation.

There are two commonly encountered pipe operators in R:

1.The native R pipe: |>
1. The magrittr pipe: %>%

For example:
```{r}
x <- c(1:10)

sqrt(x)

sum(sqrt(x))

sqrt(x) |> 
  sum()

```

The pipe makes it easier to read a sequence of operations from left to right.

# The Tidyverse   

The tidyverse is a collection of R packages designed to make data analysis easier and more consistent.

One of the most commonly used packages in the tidyverse is dplyr, which provides functions for manipulating data.

Some important dplyr functions are:

- select() — select columns;
- filter() — select rows;
- arrange() — reorder rows;
- mutate() — create or modify columns;
- rename() — rename columns;
- group_by() — define groups;
- summarise() — calculate summaries.



## Selecting Columns with select()

The select() function is used to select columns from a data frame.

First, we can look at the column names:

```{r}
colnames(dat)

dat |> 
  select(yield)

```


### Select Helpers

The select() function can also use helper functions.

  - starts_with()
  - ends_with()
  - contains(): 
  - matches(): 
  - num_range()
        
```{r}

#Select variables whose names start with a specific character or string.
dat |> select(starts_with("y"))

# Select variables whose names end with a specific character or string.
dat |> select(ends_with("loc"))

#Select variables whose names contain a specific character or string.
dat |> select(contains("e"))

#matches() uses a regular expression to identify column names.
#For example, to select variables whose names contain either "y" or "e":

dat |> select(matches("y|e"))


#num_range() is useful when variables follow a numerical naming pattern.
#for example, suppose we have variables named G18, G19, and G20:

dat_gen <- dat |> 
  mutate(value = 1) |> 
  pivot_wider(
    names_from = gen,
    values_from = value,
    values_fill = 0
)

dat_gen |> select(num_range("G", 18:20))

```


### Selecting Columns Without Tidyverse

The same operation can be performed using base R.

```{r}
dat2 <- dat[, c("env", "loc")]

#We can also select columns by their position:
dat2 <- dat[, c(3, 6)]
```

Both approaches work, but select() is often easier to read when working with more complicated data manipulation pipelines.



"filter" function: select rows that satisfy a logic test
      - between()
```{r}
# Select the trials in which 500 <= yield <= 700

# Option 1: "between" function
dat |> 
  filter(between(yield, 500, 700))

# Option 2: logical operations
dat |> 
  filter(yield >= 500 & yield <= 700)


# Other common filter patterns:

# Single condition: one location
dat |> 
  filter(loc == "Kobo")

# Multiple values: keep rows matching any of several genotypes ("%in%" instead of chaining "|")
dat |> 
  filter(gen %in% c("G01", "G02", "G03"))

# Combining conditions with "&" (AND) and "|" (OR)
dat |> 
  filter(loc == "Kobo" & year == 2001)

dat |> 
  filter(loc == "Kobo" | loc == "Melkassa")

# Negation with "!"
dat |> 
  filter(!loc == "Kobo")
```


"arrange" function: orders the rows of a data frame
```{r}
# Reorder the data frame rows according to the highest-to-low yield order

dat |> 
  arrange(desc(yield))

# Equivalent: arrange() sorts ascending by default, "-" flips it to descending
dat |> 
  arrange(-yield)

# Sorting ascending, for comparison
dat |> 
  arrange(yield)

# Sorting by multiple columns: location, then highest yield within each location
dat |> 
  arrange(loc, desc(yield))
```


"mutate" function: create new columns
```{r}
# Convert yield from kg/ha to t/ha
dat |> 
  mutate(yield_t = yield / 1000)

# Create several columns at once, and reference a column created earlier in the same mutate()
dat |> 
  mutate(
    yield_t = yield / 1000,
    yield_t_round = round(yield_t, 2)
  )

# Combine with group_by() to create a column relative to a group,
# e.g. each trial's yield deviation from its location's mean yield
dat |> 
  group_by(loc) |> 
  mutate(
    loc_mean_yield = mean(yield, na.rm = TRUE),
    yield_dev = yield - loc_mean_yield
  ) |> 
  ungroup()

# mutate() + case_when() to create a categorical column from a numeric one
dat |> 
  mutate(
    yield_class = case_when(
      yield < 500  ~ "low",
      yield < 1000 ~ "medium",
      TRUE         ~ "high"
    )
  )
```


"rename" function: change the name of a variable (column)
```{r}
# Change the name of gen and loc so that the new names were genotype and location, respectively.
# rename(new_name = old_name)

dat |> 
  rename(genotype = gen, location = loc)

# rename_with(): rename many columns at once with a function, e.g. make everything uppercase
dat |> 
  rename_with(toupper)

# rename_with() + a helper: rename only columns matching a condition
dat |> 
  rename_with(toupper, .cols = c(gen, loc))
```


"across" function: select the columns to which you want to apply a function 
```{r}
# Coerce the year variable from type numeric (double) to type character
dat |> 
  mutate(across(year, as.character)) |> 
  str()

# across() with a tidyselect helper: apply the same function to every character/factor column
dat |> 
  mutate(across(where(is.factor), as.character)) |> 
  str()

# across() with multiple named columns
dat |> 
  mutate(across(c(gen, env, loc), as.character)) |> 
  str()

```


# Group by and summarise

```{r}

dat |> 
  group_by(loc) |> 
  summarise(mean_loc = mean(yield),
            min_yield = min(yield),
            max_yield = max(yield))


```




